\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-12-25 nobuild/jss_revision/sparseHessianFD.Rnw
proof of sparsity with complex step method

When $f(x)$ is complex analytic, we can approximate $\parD{}{}{f(x)}$
and $\hess{}{f(x)}$ using the complex step
method. To do this, we extend our defintion of a real $x\in\Real{M}$
to be a complex $x\in\Complex{M}$ for which the imaginary part is
zero. We then take a Taylor
series expansion of $f(x)$ in the imaginary direction of the complex
plane \citep{SquireTrapp1998}, rearrange terms, and take the imaginary
parts of both sides.
\begin{align}
  \label{eq:complexTaylor}
  f(x+i\delta)&=f(x) + i\delta\parD{}{}{f(x)}+\mathcal{O}(\delta^2)\\
  \parD{}{}{f(x)}&\approx\frac{\Imag(f(x+i\delta))}{\delta}
\end{align}

Estimating a first derivative using the complex step method does not
require a differencing operation, but a second-order approximation of
the Hessian does \citep{AbreuStich2013}.  However, when the gradient
is provided, we can estimate the Hessian as if it were a Jacobian of
the vector-valued gradient function \citep{LaiCrassidis2008}.

@@

If this matrix were dense, we would need two evaluations of the
complex $\parD{}{}f(x)$ to estimate it.  But what if it had the same sparsity
pattern as the Hessian in Equation~\ref{eq:FDhess2sp}? First, we need
to make an assumption that structural zeros remain zero for all
complex $x\in\Complex{M}$.  Next, subtract
$\Imag(\parD{}{}{f(x_1,x_2)})$ from each column of
$\hess{}{f(x)}$. When $x$ is real, the imaginary part of the gradient
is zero, so this operation has no effect on the value of the
Hessian. But for all complex $x$, the sparsity constraints ensure that
the following identities hold:

@@

As with the finite difference method, because Equation~\ref{eq:I1c} holds
for $x_1$, it must also hold for $x_1+i\delta$, because Equation~\ref{eq:I1d} holds
for $x_2$, it must also hold for $x_2+i\delta$. Thus, for real $x$,

@@

Next, define $\Mat{Y}\in\Real{M\times C}$ such that each column is a
difference in gradients.
\begin{align}
  \label{eq:Yg}
  \Mat{Y}_c&=\nabla f(x+G_c)-\nabla f(x)
\end{align}

If $C=M$, then $\Mat{G}$ is a diagonal matrix with $\delta$ in each
diagonal element.  The matrix equation $\hess{}{}\Mat{G}=\Mat{Y}$ represents the linear approximation
$\hess{im}{}\delta\approx y_{im}$, and we can solve for all elements of $\hess{}{}$
just by computing $Y$. But if $C<M$,
there must be at least one $\Mat{G}_c$ with $\delta$ in at least two
rows. The corresponding column $\Mat{Y}_c$ is computed by perturbing multiple variables at
once, so we cannot solve for any $\hess{im}{}$
without further constraints.

@@

\subsection[Complex step method]{Complex step method}\label{sec:complex}

The \pkg{sparseHessianFD} package also allows for
estimation of the Hessian using the complex step method
\citep{SquireTrapp1998} as an alternative to finite differencing. In
this method, the definition of $\Mat{Y}_c$ in Equation~\ref{eq:Yg} is replaced with
\begin{align}
  \label{eq:complexYg}
  \Mat{Y}_c&=\Imag(\nabla f(x+iG_c))
\end{align}
The advantage of this method is that there is no longer a
subtraction operation to generate truncation error. Thus, $\delta$ can
be made arbitrarily small, and the resulting estimate of the Hessian
is more accurate.

>>>>

We are interested in estimating the Hessian only for real
$x\in\Real{M}$. However, when $f(x)$ is complex-valued over a complex domain $x\in\Complex{M}$, we can approximate $\parD{}{}{f(x)}$
and $\hess{}{f(x)}$ for real $x$ using the complex step
method.  This method comes from a Taylor
series expansion of $f(x)$ in the imaginary direction of the complex
plane \citep{SquireTrapp1998}.  After rearranging terms, and taking the imaginary
parts of both sides,
\begin{align}
  \label{eq:complexTaylor}
  f(x+i\delta)&=f(x) + i\delta\parD{}{}{f(x)}+\mathcal{O}(\delta^2)\\
  \parD{}{}{f(x)}&\approx\frac{\Imag(f(x+i\delta))}{\delta}
\end{align}

Estimating a first derivative using the complex step method does not
require a differencing operation, so there is no loss of precision
from rounding error, and the approximation of the derivative can be
made arbitrarily precise as $\delta\rightarrow 0$.  This is not the
case for second-order approximations of
the Hessian \citep{AbreuStich2013}.  However, when the gradient can be
computed exactly, we can compute a first-order approximation to
Hessian by treating it as the Jacobian of
a vector-valued function \citep{LaiCrassidis2008}.

@@

If this matrix were dense, we would need two evaluations of the
complex $\parD{}{}f(x)$ to estimate it.  But what if it had the same sparsity
pattern as the Hessian in Equation~\ref{eq:FDhess2sp}? First, we
assume that structural zeros remain zero for all
complex $x\in\Complex{M}$.  Next, subtract
$\Imag(\parD{}{}{f(x_1,x_2)})$ from each column of
$\hess{}{f(x)}$. When $x$ is real, the imaginary part of the gradient
is zero, so this operation has no effect on the value of the
Hessian. But the sparsity constraints ensure that
the following identities hold:

@@

As with the finite difference method, because Equation~\ref{eq:I1c} holds
for $x_1$, it must also hold for $x_1+i\delta$, and because Equation~\ref{eq:I2c} holds
for $x_2$, it must also hold for $x_2+i\delta$. Thus,

@@

Next, define $\Mat{Y}\in\Real{M\times C}$ such that each column is
either a difference in gradients, or the imaginary part of a
complex-valued gradient, depending on the chosen method.
\begin{align}
  \label{eq:Yg}
  \Mat{Y}_c&=
           \begin{cases}  
             \nabla f(x+G_c)-\nabla f(x)&\text{finite difference method}\\
             \Imag(\nabla f(x+iG_c))&\text{complex step method}
             \end{cases}
\end{align}

If $C=M$, then $\Mat{G}$ is a diagonal matrix with $\delta$ in each
diagonal element.  The matrix equation $\hess{}{}\Mat{G}=\Mat{Y}$ represents the linear approximation
$\hess{im}{}\delta\approx y_{im}$, and we can solve for all elements of $\hess{}{}$
just by computing $\Mat{Y}$. But if $C<M$,
there must be at least one $\Mat{G}_c$ with $\delta$ in at least two
rows. The corresponding column $\Mat{Y}_c$ is computed by perturbing multiple variables at
once, so we cannot solve for any $\hess{im}{}$
without further constraints.

@@
